A Comparison of Actor Critic and Policy Gradient \\ Algorithms for Learning Mechanical Designs

### 2022-04-29 Paper/PaperForReview.tex
few changes
final version

\listoftodos

@@

As such, this paper explores the use of RL for concurrent design of flexible-legged locomotive systems. A simplified flexible jumping system was used where, for the initial work, the control input was held fixed so that an RL algorithm was tasked with only learning optimized mechanical parameters. The rest of the paper is organized such that in the next section, similar work will be discussed. In Section \ref{sec:monopode_model}, the monopode environment details will be defined. Next, in Section~\ref{sec:control_input_exp}, the input used during training will be explained. Then, in Section~\ref{sec:learning_mech_params}, the algorithm used along with the method of the experiments will be presented. The performance of the learned designs are shown in Section~\ref{sec:results}.

@@

The training hyperparameters were selected based on the algorithm author recommendations and StableBaselines3 experimental findings. The vales are displayed in Table~\ref{tab:training_hyperameters}. All of the hyperparameters, with the exception of the rollout (Learning Starts) and the replay buffer, were set according to StableBaselines3 standards. The rollout setting was defined such that the agent could search the design space at random, filling the replay buffer with enough experience to prevent the agent from converging to a design space that was not optimal. The replay buffer was sized proportional to the number of training steps due to system memory constraints.  
%
\begin{table}[t]
        \caption{TD3 Training Hyperparameters}
        \vspace{-4mm}
        \label{tab:training_hyperameters}
        \begin{center}
        \begin{tabular}{|c||c|}
        \hline
        Hyperparameter                     & Value                           \\
        \hline
        Learning Rate, $\alpha$         & 0.001                           \\
        Learning Starts                 & 10 Steps                        \\
        Batch Size                      & 100 Transitions                 \\
        Tau, $\tau$                     & 0.005                           \\
        Gamma, $\gamma$                 & 0.99                            \\
        Training Frequency              & 1:Episode                       \\
        Gradient Steps                  & $\propto$ Training Frequency    \\
        Action Noise,  $\epsilon$       & None                            \\
        Policy Delay                    & 1 : 2 Q-Function Updates        \\
        Target Policy Noise, $\epsilon$ & 0.2                             \\
        Target Policy Clip, $c$         & 0.5                             \\
        Seed                            & 10 Random Seeds                 \\
        \hline
        \end{tabular}
        \end{center}
        \vspace{-5mm}
\end{table}
%
%
\begin{table}[t]
        \caption{PPO Training Hyperparameters}
        \vspace{-4mm}
        \label{tab:training_hyperameters_PPO}
        \begin{center}
        \begin{tabular}{|c||c|}
        \hline
        Hyperparameter                  & Value                           \\
        \hline
        Learning Rate, $\alpha$         & 0.001                           \\
        Batch Size                      & 64                              \\
        Clip Range, $\epsilon $         & 0.2                             \\
        Gamma, $\gamma$                 & 0.99                            \\
        GAE-Lambda,                     & 0.95                            \\
        Training Frequency              & 1:Episode                       \\
        Gradient Steps                  & $\propto$ Training Frequency    \\
        Policy Delay                    & 1 : 2 Q-Function Updates        \\
        Target Policy Noise, $\epsilon$ & 0.2                             \\
        Target Policy Clip, $c$         & 0.5                             \\
        Seed                            & 10 Random Seeds                 \\
        \hline
        \end{tabular}
        \end{center}
        \vspace{-5mm}
\end{table}
%

The training hyperparameters for PPO found in Table~\ref{tab:training_hyperameters_PPO} were set similarly to TD3's. All the PPO hyperameters, other than learning rate and N-steps, were set according to the algorithm's author and StableBaselines3's recommendations.
The learning rate and N-steps were adjusted as the algorithm was not able to converge within 1000 episodes given the initial hyperparameter values of 0.0003 and 2048. The learning rate and N-steps parameters were selected such that results converged in a similar amount of
episodes as TD3.

@@

The reward for the second case was:
%
\begin{equation}
        \mathbb{R}_2 = - \frac{\left (\textbf{X}  \right )_{\text{max}} - x_{s}}{x_{s}}
\end{equation}
%
where $x_s$ was the desired jump height, which was set to 0.08~m. The second case was utilized to test RL's ability to find a design that minimized the error between the maximum height reached and the desired maximum height to reach.

@@

Figure~\ref{fig:rew_vs_step} shows the rewards the agents received during training. 
At 1000 episodes the mean rewards of both algorithms when learning designs for maximum height are similar. Both have rewards within one standard deviation of the other algorithm's performance. However, PPO shows to be the preferable algorithm in this case, as it has a significantly smaller standard deviation.

This means that fewer agents would need to be trained using PPO, as training an agent using the PPO algorithm will more consistently lead to the algorithms optimal result. PPO is also the preferred algorithm when learning designs that would jump the monopode to a specified height. Figure~\ref{fig:rew_vs_step} shows that the agents trained with the PPO algorithm were able to obtain a higher maximum reward than the algorithms trained with TD3.

@@

The average and standard deviation of the spring constant and damping ratio design parameters the agents selected during training are shown in Figures~\ref{fig:spring_vs_step} and~\ref{fig:zeta_vs_step}. These figures show that the PPO and TD3 Algorithm on average converge to a similar result when learning a design to maximize jump height. However, the variation in the standard deviation of the parameter choices between the PPO agents, show that each agent's parameter choice varied greatly between episodes. This implies a higher preference for exploration in the PPO agents compared to the TD3 agents. The larger standard deviation of the TD3 algorithms does not demonstrate the algorithm's preference for exploration, instead it shows that agents trained with TD3 algorithm converged on solutions with more significant differences.

When the agents learned designs to jump the monopode to a specific height, the two algorithm's agents converged on different solutions. Additionally, the standard deviation of the spring constant learned by both algorithm's agents, and the damping ratio learned by the PPO algorithm agents, are significantly larger than when the agents were trained for maximum height. This is due to there being a wider range of possible solutions to achieve the specified height of 0.08m.
%
\begin{figure}[tb]
\begin{center}
        \includegraphics[width = 3in]{figures/ppo_vs_td3/avg_spring_comp.png}
        \caption{Spring Constant Selected During Training}
        \label{fig:spring_vs_step}
\end{center}
% \vspace{-4mm}
\end{figure}
%
\begin{figure}[tb]
        \begin{center}
                \includegraphics[width = 3in]{figures/ppo_vs_td3/avg_zeta_comp.png}
                \caption{Damping Ratio Selected During Training}
                \label{fig:zeta_vs_step}
        \end{center}
        % \vspace{-4mm}
\end{figure}

@@

\section{Future Work}
\todo[inline]{Maybe there should be a small paragraph on future work here.}

@@

\begin{figure}[tb]
\begin{center}
        \includegraphics[width=3in]{figures/comp_learning_rate/avg_rew_td3.png}
        \caption{Reward Due to Changes in TD3 Learning Rate}
        \label{fig:td3_rew}
\end{center}
% \vspace{-4mm}
\end{figure}

In the process of this project the effects of learning rate on the PPO and TD3 Algorithms were explored.
Figures~\ref{fig:ppo_rew} and \ref{fig:td3_rew} show that the Time Steps needed for the PPO algorithm to converge is very sensitive to the choice in learning rate, it also shows that the overall mean and standard deviation of the rewards is sensitive. On the other hand, TD3 does not appear to be as sensitive to changes to the learning rate, at least not in the range of learning rates explored in this study.

@@

In addition to learning rate the effects of the update rate on PPO was explored. This study found an update rate of one update per 50 time steps to be optimum. The results shown in Figure~\ref{fig:n_steps} show that an update rate of 100 is to slow as the algorithm is unable to converge by the 1000th Time Step. The figure also shows that updating every 10 steps is also a poor choice. This is because the algorithm is forced to make updates on small batch sizes.

>>>>

@@

As such, this paper explores the use of RL for concurrent design of flexible-legged locomotive systems. For this initial work, RL was evaluated for defining mechanical design parameters given a static control input. In other words, RL was evaluated as the mechanical designer in a concurrent design architecture. Two algorithms were evaluated against each other to evaluate the performance differences between actor-critic and policy gradient based algorithms. In this work, the algorithms used were Twin Delayed Deep Deterministic Policy Gradient (TD3) and Proximal Policy Optimization (PPO).

A simplified flexible jumping system was used where, for this initial work, the control input was held fixed so that an RL algorithm was tasked with only learning optimized mechanical parameters. The rest of the paper is organized such that in the next section, similar work will be discussed. In Section \ref{sec:monopode_model}, the monopode environment details will be defined. Next, in Section~\ref{sec:control_input_exp}, the input used during training will be explained. Then, in Section~\ref{sec:learning_mech_params}, the algorithms used along with the method of the experiments will be presented. The performance of the learned designs are shown in Section~\ref{sec:results} and conclusive remarks will be made in Section~\ref*{sec:conclusion}.

@@

The training hyperparameters were selected based on the algorithm author recommendations and StableBaselines3 experimental findings. The vales are displayed in Table~\ref{tab:training_hyperameters}. All of the hyperparameters, with the exception of the rollout (Learning Starts) and the replay buffer, were set according to StableBaselines3 standards. The rollout setting was defined such that the agent could search the design space at random, filling the replay buffer with enough experience to prevent the agent from converging to a design space that was not optimal. The replay buffer was sized proportional to the number of training steps due to system memory constraints.  
%
\begin{table}[t]
        \caption{TD3 Training Hyperparameters}
        \vspace{-4mm}
        \label{tab:training_hyperameters}
        \begin{center}
        \begin{tabular}{|c||c|}
        \hline
        Hyperparameter                     & Value                           \\
        \hline
        Learning Rate, $\alpha$         & 0.001                            \\
        Learning Starts                 & 10 Steps                        \\
        Batch Size                      & 100 Transitions                 \\
        Tau, $\tau$                     & 0.005                           \\
        Gamma, $\gamma$                 & 0.99                            \\
        Training Frequency              & 1:Episode                       \\
        Gradient Steps                  & $\propto$ Training Frequency    \\
        Action Noise,  $\epsilon$       & None                            \\
        Policy Delay                    & 1 : 2 Q-Function Updates        \\
        Target Policy Noise, $\epsilon$ & 0.2                             \\
        Target Policy Clip, $c$         & 0.5                             \\
        Seed                            & 10 Random Seeds                 \\
        \hline
        \end{tabular}
        \end{center}
        \vspace{-5mm}
\end{table}
%
%
\begin{table}[t]
        \caption{PPO Training Hyperparameters}
        \vspace{-4mm}
        \label{tab:training_hyperameters_PPO}
        \begin{center}
        \begin{tabular}{|c||c|}
        \hline
        Hyperparameter                  & Value                           \\
        \hline
        Learning Rate, $\alpha$         & 0.02                            \\
        N Steps                         & 50                              \\
        Batch Size                      & 64                              \\
        N Epochs                        & 10                              \\
        Gamma, $\gamma$                 & 0.99                            \\
        GAE-Lambda,                     & 0.95                            \\
        Clip Range, $\epsilon $         & 0.2                             \\
        Seed                            & 10 Random Seeds                 \\
        \hline
        \end{tabular}
        \end{center}
        \vspace{-5mm}
\end{table}
%

The training hyperparameters for PPO found in Table~\ref{tab:training_hyperameters_PPO} were set similarly to those of TD3. All the PPO hyperameters, other than learning rate and N-steps, were set according to the author of the algorithms and recommendations the recommendations of \textbf{StableBaselines3}.
The learning rate and N-steps were adjusted as the algorithm was not able to converge within 1000 episodes given the initial hyperparameter values of 0.0003 and 2048. The learning rate and N-steps parameters were selected such that results converged in a similar amount of
episodes as TD3.

@@

The reward for the second case was:
%
\begin{equation}
        \mathbb{R}_2 = - \frac{\left (\textbf{X}  \right )_{\text{max}} - x_{s}}{x_{s}}
\end{equation}
%
where $x_s$ was the desired jump height, which was set to 0.08~m. The second case was utilized to test the ability of RL to find a design that minimized the error between the maximum height reached and the desired maximum height to reach.

@@

Figure~\ref{fig:rew_vs_step} shows the rewards the agents received during training. 
At 1000 episodes the mean rewards of both algorithms when learning designs for maximum height are similar. Both have rewards within one standard deviation of the performance of the other algorithms. However, PPO shows to be the preferable algorithm in this case, as it has a significantly smaller standard deviation.
This means that fewer agents would need to be trained using PPO, as training an agent using the PPO algorithm will more consistently lead to the algorithms optimal result. PPO is also the preferred algorithm when learning designs that would jump the monopode to a specified height. Figure~\ref{fig:rew_vs_step} shows that the agents trained with the PPO algorithm were able to obtain a higher maximum reward than the algorithms trained with TD3.

@@

The average and standard deviation of the spring constant and damping ratio design parameters the agents selected during training are shown in Figures~\ref{fig:spring_vs_step} and~\ref{fig:zeta_vs_step}. These figures show that the PPO and TD3 Algorithm on average converge to a similar result when learning a design to maximize jump height. However, the variation in the standard deviation of the parameter choices between the PPO agents, show that each agent's parameter choice varied greatly between episodes. This implies a higher preference for exploration in the PPO agents compared to the TD3 agents. The larger standard deviation of the TD3 algorithm does not demonstrate a preference for exploration, instead it shows that agents trained with TD3 algorithm converged on solutions with more significant differences.

When the agents learned designs to jump the monopode to a specific height, the two agents for both algorithms converged on different solutions. Additionally, the standard deviation of the spring constant learned by both agents for both algorithms, and the damping ratio learned by the PPO algorithm agents, are significantly larger than when the agents were trained for maximum height. This is due to there being a wider range of possible solutions to achieve the specified height of 0.08m.
%
\begin{figure}[tb]
\begin{center}
        \includegraphics[width = 3in]{figures/ppo_vs_td3/avg_spring_comp.png}
        \caption{Spring Constant Selected During Training}
        \label{fig:spring_vs_step}
\end{center}
% \vspace{-4mm}
\end{figure}
%
\begin{figure}[tb]
        \begin{center}
                \includegraphics[width = 3in]{figures/ppo_vs_td3/avg_zeta_comp.png}
                \caption{Damping Ratio Selected During Training}
                \label{fig:zeta_vs_step}
        \end{center}
        % \vspace{-4mm}
\end{figure}

@@

\section{Future Work}
Future work should include comparing the performance of the algorithms with increasingly complex systems. This work focused on optimizing a limited set of parameters for the monopode jumping system, which is a simplified flexible jumping system, however the utilization of RL for mechanical design could be expanded into many different types of systems. Additionally, as was mentioned in Section~\ref*{sec:intro}, the utilization of RL as a mechanical design optimizer should be evaluated in a concurrent design architecture. As was previously mentioned, finding optimal control strategies is often a challenge given a system design, therefore developing an architecture that could concurrently optimize both aspects of a robotics system, is of interest.

@@

%  !!!!!!!!!!!!!!!! !!!!!!!!!!!! !!!!!!!!!!!!! !!!!!!!!!!!!!! 
% THIS IS NOT THE RIGHT DATA. THE TRAINING SCRIPT DID NOT UPDATE LEARNING RATE.

% \begin{figure}[tb]
% \begin{center}
%         \includegraphics[width=3in]{figures/comp_learning_rate/avg_rew_td3.png}
%         \caption{Reward Due to Changes in TD3 Learning Rate}
%         \label{fig:td3_rew}
% \end{center}
% % \vspace{-4mm}
% \end{figure}

In the process of this project the effects of learning rate on the PPO algorithm was explored to find a value that would allow PPO to perform in a similar manner to TD3. Figure~\ref{fig:ppo_rew} shows that the time steps needed for the PPO algorithm to converge is very sensitive to the choice in learning rate, it also shows that the overall mean and standard deviation of the rewards is sensitive as well.

@@

In addition to learning rate the effects of the update rate on PPO was explored. This study found an update rate of one update per 50 time steps to be optimum. The results shown in Figure~\ref{fig:n_steps} show that an update rate of 100 is to slow as the algorithm is unable to converge by the 1000th time step. The figure also shows that updating every 10 steps is also a poor choice. This is because the algorithm is forced to make updates on small batch sizes.
